Using \LaTeX\ to submit manuscripts to peer reviewed journals.

### 2014-02-17 docs/thesis/chap-cern_math_explorer.tex
more work on document

\chapter{CERN Math Explorer}
\label{chapter-cern_math_explorer}
In this chapter we propose our solution for a scalable math based oriented search engine. We present the key points of our system, and how they are related to previous built systems.

\section{Information Retrieval Model}
Our main goal is to develop a system that compliments the already available search capabilities in Invenio and in particular the CDS instance. Our system was developed using the Apache Solr/Lucene framework.

@@

\subsubsection{Numeric Approximation} 
Some equations relate specific numeric quantities. While most of the previously explorer systems, handled <mn> tags by only representing them as a constant identifier, this approach is not suitable. A lot of articles in CDS relate to specific experiments under certain conditions, and require a high level of precision in the measurements.

\subsection{Structural Features}
\subsubsection{}
\subsubsection{Algebraic structures}
This set of structures consist on purely algebraic relationships between the expressions.
Inspired by predefined integral tables for different kinds of expressions, we decided to extract from a given equation, whether it has the occurrence of these kind of patterns.
An example of this type of structures would be: $X\textunderscore*(X\textunderscore + A\textunderscore)$ where $X\textunderscore$ and $A\textunderscore$ can be any type of expression. Sample equations matching this pattern can be $C (C+1)$ or $\frac{(\sin{\theta}\cos{\theta})^2}{2a} (\frac{(\sin{\theta}\cos{\theta})^2}{2a} + \tan{\frac{\theta+\phi}{2}})$. This structural analysis of equations is more powerful than the unification of variables and constants proposed in previously discussed projects because these patterns, as shown in the second example, can match an arbitrary complex expression. The drawback of our solution is the need to manually select representative and useful patterns. This selection was done based on standard based tables of integrals similar to the ones that Computational Algebraic Systems employ to compute them.

>>>>

\chapter{MASE(CDS)}
\label{chapter-cern_math_explorer}
In this chapter we present our solution called MASE(CDS), Math Aware Search Engine (For CDS). We present the key points of our system, and how they are related to previous built systems.

\section{Features Extraction}
Our main goal is to develop a system that compliments the already available search capabilities in Invenio and in particular the CDS instance. Our system was developed on top of the  the Apache Solr/Lucene framework. CDS already uses Solr as a search engine layer, so developing MASE using this technology supposed a reasonable choice. One of the main steps while building an IR tool, is understanding the data and what kind of features can be extracted from it: If the system is based on text, then tasks as stemming, handling of synonyms, misspelling and so on, such be taken into consideration; if the system being developed is based on images, then examples of features can be histograms of colors, geometrical features and so on. During the literature review, we discovered that most of the projects handled the notational aspects of the expressions, that is, the real naming of the terms in a given equation and each project developed its own set of heuristics to handle the structural component.

@@

\subsubsection{Numeric Approximation} 
Some equations relate specific numeric quantities. While most of the previously explorer systems, handled <mn> tags by only representing them as a constant identifier, this approach is not suitable. A lot of articles in CDS relate to specific experiments under certain conditions, and require a high level of precision in the measurements. However, different experiments of the same phenomenon, can result in different measurements and we wanted to include this variability in our system. To handle this, when a token with a certain numeric quantity is discovered, we apply iterative rounding and index all the different approximations in the same position as the original element. For example the token <mn>2.0135</mn> will produce the following sequence of tokens: <mn>2.0135</mn><mn>2.014</mn><mn>2.01</mn><mn>2.0</mn>.

\subsection{Structural Features}

\subsubsection{Algebraic structures}
The first set of structures consist on purely algebraic relationships between the expressions.
Inspired by predefined integral tables for different kinds of expressions, we decided to extract from a given equation, whether it has the occurrence of these kind of patterns.
An example of this type of structures would be: $X\textunderscore*(X\textunderscore + A\textunderscore)$ where $X\textunderscore$ and $A\textunderscore$ can be any type of expression. Sample equations matching this pattern can be $C (C+1)$ or $\frac{(\sin{\theta}\cos{\theta})^2}{2a} (\frac{(\sin{\theta}\cos{\theta})^2}{2a} + \tan{\frac{\theta+\phi}{2}})$. This structural analysis of equations is more powerful than the unification of variables and constants proposed in previously discussed projects because these patterns, as shown in the second example, can match an arbitrary complex expression. The drawback of our solution is the need to manually select representative and useful patterns. This selection was done based on standard based tables of integrals similar to the ones that Computational Algebraic Systems employ to compute them.

\subsubsection{Non-formal mathematical variations}
This set of structures represent common operations that cat appear in a given expression, but its identification is based on notation factors. A simple example of this can be the matrix decomposition $A = VDV^{-1}$ where the actual name of the variables $A$ and $V$ is not meaningful but the real importance is the fact that a given variable appears alongside its inverse in a matrix multiplication form. We would like to assign a partial match to any other kind of this decomposition say $B = MRM^{-1}$. these types of structures live in-between the notational and structural categories but, as it will be discussed, we identify them using Mathematica and encompass them as structural features.

### 2014-02-17 docs/thesis/chap-evaluation.tex
more work on document

>>>>

\section{Results quality}
In this section we are interested in evaluating the quality of the results provided by our system.
The type of queries that MASE was designed for, are typical equations that can be found in a scientific publication. 
Out main user case is that a user of CDS is interested in finding documents inside the collection whose equations are related to the one that is provided.

\subsection{Equations}
The evaluated queries were provided by physics experts working at CERN from their own research topics. The queries were provided in \LaTeX format as the users used them in their own publications.
Table \ref{equations_table} presents the set of queries that were evaluated, and a simple explanation about it.

\begin{longtable}{|m{1cm}|>
{\centering\arraybackslash}m{3cm}|>
{\centering\arraybackslash}m{5cm}|>
{\centering\arraybackslash}m{5cm}|
}
\hline 
\textbf{Id} & 
\textbf{Description} &
\textbf{MathML code} &
\textbf{Visual Rendering}

\\
\hline
Eq1 & Time-independent Schr\char"00F6dinger equation & \scriptsize {\codefont <math><mfenced close=")" open="("> <mrow> <mo>-</mo> <mfrac> <msup> <mi>ℏ</mi> <mn>2</mn> </msup> <mrow> <mn>2</mn> <msub> <mi>m</mi> <mi>e</mi> </msub> </mrow> </mfrac> <msup> <mo>∇</mo> <mn>2</mn> </msup> <mo>+</mo> <mi>V</mi> <mfenced close=")" open="("> <mi>r</mi> </mfenced> </mrow> </mfenced> <mi>ψ</mi> <mfenced close=")" open="("> <mi>r</mi> </mfenced> <mo>==</mo> <mi>E</mi> <mi>ψ</mi> <mfenced close=")" open="("> <mi>r</mi> </mfenced> </math>} & $\left (-\frac{\hbar^{2}}{2m_e}\nabla^{2} + V(\mathbf{r})\right )\psi(\mathbf{r}) = E\psi(\mathbf{r})$ \\ \hline

\caption{Software used during evaluations}
\label{equations_table}
\end{longtable}

\section{Querying Performance}

### 2014-02-17 docs/thesis/chap-related_work.tex
visio drawing
more work on document

\subsection{DLMFSearch}
\subsection{MathWebSearch}
\subsection{LaTEXSearch}
LaTEXSearch \url{http://latexsearch.com/} is a system developed by the scientific publisher Springer that allows to search over a database of around eight million latex snippets extracted from their publications. The system, unfortunately, is proprietary and no details are provided about the internals of the system.

>>>>

\subsection{DLMFSearch}
\subsection{MathWebSearch}
MathWebSearch\cite{mathwebsearch} (Currently on its 0.5 version) is a open-source search engine for mathematical equations. While most of the documentation relates on the architecture of the system and how they address scalability; the indexing technique is also very interesting. The system implements and idea proposed by Peter Graf in \cite{substitution_tree_indexing} called substitution tree indexing. This idea can be viewed as a generalization of the variable and constant unification employed by MIaS

\subsection{LaTEXSearch}
LaTEXSearch \url{http://latexsearch.com/} is a system developed by the scientific publisher Springer that allows to search over a database of around eight million latex snippets extracted from their publications. The system, unfortunately, is proprietary and no details are provided about the internals of the system.

### 2014-02-17 docs/thesis/chap-theoretical_background.tex
more work on document

\section{Vector Space Model}

\section{Probabilistic Relevance Model}
The probabilistic relevance model was developed in 1976 by Robertson and Jones\cite{probabilistic_relevance_model}. The similarity of a document $d_{j}$ to a query $q$ is given by the probability of $d_{j}$ being relevant to $q$. It assumes there is a set $R$ that is preferred to be the answer set for $q$. The model assumes that the document in $R$ are relevant to $q$ and the documents in $\bar{R}$ are not relevant. $R$ then is the set that maximizes:

$$sim(d_{j},q) = \frac{ P(R|d_{j})} {P(\bar{R} |d_{j})} $$

@@

The Okapi BM25 ranking function is one of the current state-of-the-art models in information retrieval\cite{okapibm25}. The actual scoring function is called BM25 and Okapi refers to the first system implementing this function in the 1980s. 
Given a query $Q$ as a vector of keywords $q_{1}, q_{2}, ... , q_{n}$ then the score for a document $d$ is:

$$score(d, Q) = \sum_{i=1}^{n}IDF{(q_i)}*\frac{ frec(q_i, D)}{frec(q_i, D)+k_1*((1-b)+b*\frac{|D|}{avgdl})} $$
and $$IDF(q_i) = \log{\frac{N - df(q_i)+0.5}{df(q_i)+0.5}} $$

@@

$$score(d, Q) = \sum_{i=1}^{n} \frac{weigth(q_i, d)}{k+ weigth(q_i, d)} * IDF(q_i)$$
and 
$$weight(q_i, d) = \sum_{j=1}^m \frac{frec(q_i, D, field_j)*boost_j)}{((1-b_j) + b_j*\frac{|l_j|}{avgfl_j})} $$
where $m$ is the number of different fields, $boost_j$ is the relative importance for each field, $b_j$ is analogous to the $b$ constant in the BM25 group of equations, $|l_j|$ is the length of the $field_j$ and $avgfl_j$ is the average length of the $field_j$.

>>>>

\section{Vector Space Model}
The vector space model\cite{vector_space_model} which was developed between the 1960s and 1970s alongside the SMART Information Retrieval System. In this model, documents (from a collection $D$) and queries are represented as vectors $d$ and $q$ in a vector space where each index term corresponds to a dimension. If a term $t_i$ occurs in a given query or document, its i-th  coordinate will have a non zero value. Index terms can be keywords or phrases in the context of textual search or any other type of feature that can be extracted from the elements in a given specific context.
The scoring of a document for a given query is computed as the cosine of the angle between both vectors which can be expressed as: 
$$score(d, q) = cos\theta = \frac{d \cdot q}{\parallel d \parallel \parallel q \parallel}  = \frac{\sum_{i=1}^{n}weight(d_i, d)*weight(q_i, q)}{\sqrt{\sum_{i=1}^{n}weight^2(d_i, d)}\sqrt{\sum_{i=1}^{n}weight^2(q_i, q)}} $$

Different weighting schemes can be employed, and the most standard one is called tf-idf (Term frequency - Inverse document frequency) where terms that appear less frequently in the corpus, are assigned higher values (Intuitively the term "the" will be less useful for identifying a relevant document than the term "Shakespeare" in the context of an English word based search engine). Therefore:

$$weight(t, d) = tf(t,d) idf(t)$$

There are several variations for each of the two components. Some approaches for $tf$ include the raw frequency; Boolean frequency: $tf(t,d) = 1$ if $t$ occurs in $d$ or 0 if not; logarithmically scaled: $tf(t,d) = 	\log(frequency(t,d) +1)$
and the augmented frequency: $tf(t,d) = 0.5 + \frac{0.5*frequency(t,d)}{ \max\{frequency(t^\prime,d) | t^\prime \in d\} }$. For computing the inverse document frequency the standard approach follows $idf = \log \frac{|D|}{1+\{d \in D: t \in d \}} $

The main drawbacks of this approach reside in the assumption that the terms are independent.

\section{Probabilistic Relevance Model}
The probabilistic relevance model was developed in 1976 by Robertson and Jones\cite{probabilistic_relevance_model}. The similarity of a document $d$ to a query $q$ is given by the probability of $d$ being relevant to $q$. It assumes there is a set $R$ that is preferred to be the answer set for $q$. The model assumes that the document in $R$ are relevant to $q$ and the documents in $\bar{R}$ are not relevant. $R$ then is the set that maximizes:

$$sim(d,q) = \frac{ P(R|d)} {P(\bar{R} |d)} $$

@@

The Okapi BM25 ranking function is one of the current state-of-the-art models in information retrieval\cite{okapibm25}. The actual scoring function is called BM25 and Okapi refers to the first system implementing this function in the 1980s. 
Given a query $q$ as a vector of keywords $q_{1}, q_{2}, ... , q_{n}$ then the score for a document $d$ is:

$$score(d, q) = \sum_{i=1}^{n}IDF{(q_i)}*\frac{ frec(q_i, D)}{frec(q_i, D)+k_1*((1-b)+b*\frac{|D|}{avgdl})} $$
and $$IDF(q_i) = \log{\frac{N - df(q_i)+0.5}{df(q_i)+0.5}} $$

@@

$$score(d, q) = \sum_{i=1}^{n} \frac{weigth(q_i, d)}{k+ weigth(q_i, d)} * IDF(q_i)$$
and 
$$weight(q_i, d) = \sum_{j=1}^m \frac{frec(q_i, D, field_j)*boost_j)}{((1-b_j) + b_j*\frac{|l_j|}{avgfl_j})} $$
where $m$ is the number of different fields, $boost_j$ is the relative importance for each field, $b_j$ is analogous to the $b$ constant in the BM25 group of equations, $|l_j|$ is the length of the $field_j$ and $avgfl_j$ is the average length of the $field_j$.
